---
title: "Introduction to myoRDNA"
author: "MuscleOmicsLab"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Introduction to myoRDNA}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r library}
library(myoRDNA)
```

# Introduction

Ribosomal DNA (rDNA) consists of repetitive genomic sequences
that encode ribosomal RNA (rRNA), an essential structural and
functional component of ribosomes. The organisation, copy number
and epigenetic regulation of rDNA are important determinants
of ribosome biogenesis and cellular protein synthesis.

Variation in rDNA copy number and DNA methylation has been
investigated in the context of ageing, cellular differentiation,
disease and changes in cellular metabolic activity. However,
the repetitive organisation of rDNA presents substantial
challenges for conventional sequencing and bioinformatics
approaches, particularly when estimating copy number or
analysing DNA methylation.

`myoRDNA` is an R package developed by MuscleOmicsLab
to facilitate the analysis of rDNA copy number and DNA
methylation using high-throughput sequencing data.

The package provides a collection of functions for extracting
rDNA-specific data, estimating copy number, calculating
methylation levels and identifying differential methylation
between experimental conditions.

# Package functionality

`myoRDNA` provides tools for four principal analytical tasks:

1. **rDNA data extraction and preprocessing:** Identifying
   rDNA-associated sequencing data, extracting genomic
   regions of interest and adjusting genomic coordinates.

2. **rDNA copy-number estimation:** Estimating relative
   and absolute rDNA copy number using sequencing coverage.

3. **DNA methylation analysis:** Quantifying methylation
   levels at individual CpG sites and across predefined
   rDNA regions.

4. **Differential methylation analysis:** Comparing methylation
   between experimental conditions using statistical models,
   including binomial and beta-binomial regression.

These functions are designed to support reproducible analyses
and integrate with established R and Bioconductor packages,
including `methylKit` and `GenomicRanges`.

# Installation

The development version of `myoRDNA` is available from GitHub
and can be installed using the `remotes` package.

```{r installation, eval=FALSE}
install.packages("remotes")

remotes::install_github(
  "MuscleOmicsLab/myoRDNA",
  dependencies = TRUE
)
```

Load the package using:

```{r load-package}
library(myoRDNA)
```

# Overview of the analytical workflow

A typical analysis begins with aligned sequencing data
that have undergone appropriate quality control and
preprocessing. For how to construct the adapted genomes 
to extract methylation data for rDNA please visit the 
paper by [Vaughan et al.](https://doi.org/10.64898/2026.08.13.744441)


For DNA methylation analysis, methylated and unmethylated
read counts can be imported into R using `methylKit`.
The resulting objects provide the basis for downstream
analyses implemented in `myoRDNA`. The exact preprocessing 
requirements depend on the sequencing technology, reference 
genome and experimental design.

# Loading your data into myoRDNA

This section demonstrates how to import and prepare DNA methylation
data for analysis using `myoRDNA`.

We use an example mouse whole-genome bisulfite sequencing (WGBS)
dataset comprising ten samples from two experimental groups:
six old mice (OLD) and four young mice (YOUNG). For additional 
information about the original dataset and experimental methodology, 
see the [original publication](https://doi.org/10.1016/j.isci.2025.112144). 
In this document we make no conclusions about the data, it is simply 
being used to demonstrate `myoRDNA` and the dataset has been reduced 
as a result. 

## Step 1: Load the example methylation dataset

The example data are distributed with `myoRDNA` in the
`inst/extdata` directory.

We use `system.file()` to retrieve the location of the dataset
within the installed package. This approach ensures that the
example works regardless of where the package is installed.
```{r load-data}
library(myoRDNA)
library(methylKit)

file_path <- system.file(
  "extdata",
  "filtered_meth_raw_chr1_chrRDNAm.rds",
  package = "myoRDNA",
  mustWork = TRUE
)

loaded_data <- readRDS(file_path)
```
The RDS file contains previously processed methylation data
from chromosome 1 and the rDNA chromosome (`chrRDNAm`).
The underlying data retain information about methylation
coverage, methylated read counts, unmethylated read counts
and genomic positions.

As the data have already been processed and saved, users
do not need to repeat the original sequencing alignment
and methylation extraction procedures to follow this
tutorial.

## Step 2: Define experimental conditions

The dataset contains six OLD samples and four YOUNG samples.
We define a binary treatment vector to distinguish these
experimental groups:

- `0`: OLD
- `1`: YOUNG

The order of the treatment assignments must correspond
exactly to the order of the samples in `loaded_data`.
```{r define-treatment}
treatment <- c(
  0, 0, 0, 0, 0, 0,
  1, 1, 1, 1
)

num_samples <- length(treatment)
```
This is a necessary component required for interacting with `methylKit` and 
this can be simply carried over for use in `myoRDNA`. The number of samples 
needs to be a numeric value and should be either the number of samples in the
whole dataset or across two different conditions. We also define a named list 
containing the sequencing accession numbers associated with each condition.
```{r define-samples}
samples <- list(
  'OLD' = c(
    "SRR28431752", "SRR28431753", "SRR28431754", "SRR28431755", 
    "SRR28431756", "SRR28431757"
  ),
  'YOUNG' = c(
    "SRR28431748", "SRR28431749", "SRR28431750", "SRR28431751"
  )
)
```
This is an important structure and needs to be of the same format as shown 
above for tracking samples when comparing rDNA metrics (such as copy number 
and  aggregate methylation proportion) in the later steps. Please ensure that 
the sample ids match the sample ids for the corresponding data in `methylKit`, 
otherwise output will be `NULL`.

## Step 3: Construct a methylRawList object

This step is only necessary for following along and using the `rds` data from 
the `myoRDNA`, if the user is using these features directly interfacing with the
`methylKit` object than this isn't needed.

We construct a `methylRawList` object by combining the
individual sample objects with their treatment assignments.

```{r create-methylrawlist}
stopifnot(length(loaded_data) == length(treatment))

loaded_data <- methods::new(
  "methylRawList",
  loaded_data,
  treatment = treatment
)
```

A `methylRawList` stores the methylation measurements from multiple samples 
alongside their experimental group assignments. This structure allows 
methylation data to be processed consistently across samples and enables 
subsequent comparisons between experimental conditions. 

## Step 4: Extract rDNA methylation data

The example dataset includes methylation measurements from chromosome 1 and the 
rDNA chromosome.

We use `get_rDNA_data()` to extract only the measurements associated with the 
rDNA chromosome.

```{r extract-rdna}
meth_rDNA <- get_rDNA_data(
  loaded_data,
  num_samples,
  rDNA_chr_name = "chrRDNAm"
)
```

The `rDNA_chr_name` argument specifies the chromosome identifier used in the 
adapted reference genome.

In this example, the mouse rDNA chromosome is named `chrRDNAm`. This identifier 
must match the chromosome name in the original methylation data.

The resulting `meth_rDNA` object contains the rDNA-associated methylation 
measurements for all ten samples, preserving the information required for
downstream analysis. If the name of the chromosome associated with rDNA is not 
known (in this case `chrRDNAm`) the user can make use of the identify unique
chromosomes.
```{r unique-chromosomes}
unique_names <- get_unique_chr_names(
    loaded_data,
    num_samples
)
```
This returns an array with the unique chromosome names that are included, 
additionally it performs a check to verify that all samples have consistent 
chromosome names. 

## Step 5: Adjust rDNA genomic coordinates

Depending on how the adapted reference genome was
constructed, rDNA genomic coordinates may require
adjustment to correspond to the reference sequence
used for downstream analysis.

The `shifting()` function applies a common coordinate
transformation to the selected rDNA data.

First, we specify the coordinate cutoff:

```{r define-cutoff}
cutoff <- 35000
```

We then apply the coordinate shift:

```{r shift-rdna}
meth_rDNA_shifted <- shifting(
  meth_rdna_raw = meth_rDNA,
  cutoff = cutoff,
  treatment = treatment
)
```

The function identifies the maximum genomic start position across the input 
samples and calculates the shift as

$$
s = x_{\max} - c,
$$
where $x_{\max}$ denotes the maximum genomic start position and $c$ is the 
user-defined cutoff. The transformed genomic coordinates are

$$
x_{\mathrm{shifted}} = x_{\mathrm{original}} - s.
$$
The same transformation is applied to the start and end positions of every 
retained methylation record. Records falling outside the permitted coordinate 
range are removed, and the transformed data are returned as a `methylRawList`
containing the original treatment assignments.

The cutoff of 35,000 is specific to the adapted reference used in this example. 
Users analysing data generated with a different reference should select an 
appropriate cutoff based on their reference sequence and coordinate convention. 
The shifting is specific to the human and mouse genome, not enough is known 
about rDNA contig in the rat for this shifting. 

The need for shifting is obvious when plotting the `numCs` against genomic 
position. The activity will start just after genomic position at the point $s$. 
The shifting is vital for regional analysis of the 18S for differential 
methylation. However, functions surrounding the rDNA copy number will produce an
interpretative result without shifting.

# rDNA copy-number analysis

Ribosomal DNA (rDNA) is organised into tandemly repeated units, and the number 
of these repeats can vary between individuals and experimental conditions. 
Estimating rDNA copy number provides a means of investigating variation in rDNA 
abundance across biological samples.

`myoRDNA` provides two functions for estimating rDNA copy number:

- `relative_rDNA_CN()`: Estimates relative rDNA copy number using 
  sequencing coverage.
- `absolute_rDNA_CN()`: Estimates absolute rDNA copy number by normalising rDNA 
  coverage against genomic coverage and accounting for genome ploidy.
  
These estimators have different sequencing requirements.

| Sequencing method | Relative copy number | Absolute copy number |
|:------------------|:--------------------:|:--------------------:|
| Whole-genome bisulfite sequencing (WGBS) | Yes | Yes |
| Reduced-representation bisulfite sequencing (RRBS) | Yes | No |

Absolute copy-number estimation is restricted to WGBS in `myoRDNA`, as it 
requires a genome-wide coverage reference. RRBS enriches selected genomic 
fragments, which makes its sequencing coverage unsuitable for directly 
estimating absolute genomic copy number.

Relative copy-number estimates can be obtained from either WGBS or RRBS, 
although comparisons should be restricted to samples generated using compatible 
sequencing and preprocessing protocols.

## Absolute rDNA copy number

The absolute rDNA copy-number estimator uses the ratio
of rDNA sequencing coverage to reference genomic
coverage.

For a diploid genome, the estimator is defined as

$$
\boxed{
CN_{\mathrm{absolute}} =
2 \times
\frac{\overline{C}_{\mathrm{rDNA}}}
     {\overline{C}_{\mathrm{genome}}}
}
$$

where:

- \(\overline{C}_{\mathrm{rDNA}}\) is the coverage
  estimate for the rDNA reference sequence.
- \(\overline{C}_{\mathrm{genome}}\) is the corresponding
  genomic reference coverage.
- The factor of two accounts for the two copies of
  each autosome in a diploid genome.

For a haploid genome, the estimator becomes

$$
CN_{\mathrm{absolute}} =
\frac{\overline{C}_{\mathrm{rDNA}}}
     {\overline{C}_{\mathrm{genome}}}.
$$

More generally, for a genome with reference ploidy \(P\),

$$
CN_{\mathrm{absolute}} =
P \times
\frac{\overline{C}_{\mathrm{rDNA}}}
     {\overline{C}_{\mathrm{genome}}}.
$$

The current implementation supports diploid
(\(P=2\)) and non-diploid (\(P=1\)) calculations.

### Calculating absolute copy number

We use the previously loaded mouse WGBS dataset
to demonstrate absolute rDNA copy-number estimation.

```{r absolute-copy-number}
absolute_CN <- absolute_rDNA_CN(
  Meth_object = loaded_data,
  num_samples = num_samples,
  autosomes = TRUE,
  rDNA_chr = "chrRDNAm",
  diploid = TRUE
)
```

The function accepts the following arguments:

- `Meth_object`: The methylation dataset containing
  both rDNA and genomic reference coverage.
- `num_samples`: The number of samples to analyse.
- `autosomes`: Specifies whether autosomal coverage
  should be used as the genomic reference.
- `rDNA_chr`: The chromosome identifier corresponding
  to the rDNA reference sequence.
- `diploid`: Specifies whether to multiply the
  coverage ratio by two.

For this example, `diploid = TRUE` accounts for
the diploid mouse genome, and `rDNA_chr = "chrRDNAm"`
identifies the rDNA reference in the adapted genome.

The resulting copy-number estimates can be inspected
using:

```{r inspect-absolute-copy-number}
absolute_CN
```
Additionally a readout should appear with results for inspection. These 
estimates should be interpreted with consideration of reference-genome 
composition, sequencing coverage, alignment ambiguity and potential biases 
associated with mapping reads to repetitive rDNA sequences.

## Relative rDNA copy number

Relative rDNA copy number provides a normalised measure of rDNA abundance that 
can be used to compare samples processed using compatible sequencing protocols.
Unlike absolute copy-number estimation, relative estimation is supported for 
both WGBS and RRBS data in `myoRDNA`.

The relative estimator is calculated using the ratio of summed rDNA coverage to 
summed genomic reference coverage:

$$
\boxed{
CN_{\mathrm{relative}} =
\frac{\displaystyle\sum_{i=1}^{N_r}
C_{\mathrm{rDNA},i}}
{\displaystyle\sum_{j=1}^{N_g}
C_{\mathrm{genome},j}}
}
$$

where:

- \(C_{\mathrm{rDNA},i}\) denotes the coverage of
  the \(i\)-th retained rDNA position.
- \(C_{\mathrm{genome},j}\) denotes the coverage
  of the \(j\)-th retained genomic reference position.
- \(N_r\) and \(N_g\) are the numbers of retained
  rDNA and genomic reference positions, respectively.

The numerator and denominator therefore represent the total coverage across the 
rDNA and genomic reference positions used by the estimator. Unlike the absolute 
estimator, the relative estimator does not apply a diploidy correction in the 
current implementation.

### Calculating relative copy number

The following example calculates relative rDNA copy number for the previously 
loaded mouse WGBS dataset.
```{r relative-copy-number}
relative_CN <- relative_rDNA_CN(
  Meth_object = loaded_data,
  num_samples = num_samples,
  autosomes = TRUE,
  rDNA_chr = "chrRDNAm",
  diploid = TRUE
)
```
The arguments are analogous to those used for `absolute_rDNA_CN()`. Although 
`diploid = TRUE` is accepted, the current implementation calculates the same 
relative coverage ratio regardless of this argument. The resulting estimates 
can be inspected using:
```{r inspect-relative-copy-number}
relative_CN
```
The same function can also be applied to appropriately prepared RRBS 
methylation data. However, because RRBS preferentially captures particular 
genomic fragments, its relative estimates depend on the targeted regions and 
their representation in the sequencing data. Consequently, relative copy-number 
estimates from RRBS and WGBS should not be assumed to be directly 
interchangeable.

## Considerations for copy-number estimation

Several factors should be considered when interpreting rDNA copy-number 
estimates. First, the repetitive nature of rDNA introduces potential alignment 
ambiguity. Reads derived from different rDNA repeat units will align to the 
same reference sequence.

Second, sufficient sequencing coverage is required to obtain stable estimates. 
In particular, absolute copy-number estimation depends on having a suitable 
genomic reference against which rDNA coverage can be normalised. This can 
become particularly difficult with PBAT which typically provides lower 
alignment rates.

Finally, the adapted reference genome and chromosome identifiers must be 
consistent with those used during sequencing alignment and methylation 
extraction. For reproducible analyses, users should document the reference 
genome, coverage-filtering thresholds, sequencing methodology and genomic 
regions used for normalisation.

# Regional rDNA methylation analysis

The ribosomal DNA (rDNA) repeat contains several distinct functional regions, 
including the sequences encoding ribosomal RNA and the internal and external 
transcribed spacers.

DNA methylation may vary across these regions. Consequently, analysing 
methylation across the entire rDNA repeat may obscure region-specific 
variation.

`myoRDNA` provides functions to partition rDNA methylation data into predefined 
genomic regions and calculate aggregate methylation metrics for individual 
samples and experimental conditions.

## Step 1: Partition rDNA into genomic regions

The `sort_by_rDNA_region()` function separates methylation measurements into 
their corresponding rDNA regions using predefined genomic coordinates.

The package supports the following reference assemblies:

| Species | Reference assembly |
|:--------|:-------------------|
| Mouse | `mm39` |
| Rat | `rno7` |
| Human | `hg38` |

For the example mouse WGBS dataset, we use the `mm39` reference assembly.

```{r sort-rdna-regions}
region_res <- sort_by_rDNA_region(
  meth_rDNA,
  num_samples,
  species = "mm39",
  min_cov = 10
)
```

The `species` argument specifies the reference assembly used to identify the 
rDNA regions. Users analysing rat or human data should specify `rno7` or
`hg38`, respectively.

The `min_cov` argument defines the minimum sequencing coverage required for a 
CpG position to be retained. In this example, CpG positions with fewer than ten
sequencing reads are excluded.

The resulting `region_res` object organises methylation measurements by sample 
and rDNA region, allowing downstream calculations to be performed independently
for each region.
### rDNA regions

Depending on the reference assembly, rDNA measurements
can be assigned to the following annotated regions:

- **18S:** Encodes the 18S ribosomal RNA component of the small ribosomal 
subunit.
- **5.8S:** Encodes 5.8S ribosomal RNA.
- **28S:** Encodes the 28S ribosomal RNA component of the large ribosomal 
subunit.
- **5' ETS:** The external transcribed spacer located upstream of the 18S 
region.
- **ITS1:** The internal transcribed spacer between the 18S and 5.8S regions.
- **ITS2:** The internal transcribed spacer between the 5.8S and 28S regions.
- **3' ETS:** The external transcribed spacer downstream of the 28S region.

These regions can be analysed individually to investigate spatial variation in 
rDNA methylation. It is important to note that the human and mouse have all 
regions in the list above. However, the rate is limited to the `18S`, `28S`, 
and `5.8S`.

## Step 2: Calculate aggregate methylation metrics

Following regional partitioning, the `calculate_aggregate_meth()` function can 
be used to summarise methylation measurements within each rDNA region.

For this example, we calculate aggregate methylation for the 18S region.

```{r aggregate-methylation}
region <- "18S"

aggregate_18S <- calculate_aggregate_meth(
  region_res,
  region,
  conditions = samples,
  med = FALSE
)
```
The `region` argument specifies the rDNA region to analyse, while `conditions` 
associates samples with their experimental groups.

The `med` argument controls the choice of summary statistic. In the mean-based 
calculation shown here, it is set to `FALSE`.

The resulting object contains aggregate methylation measurements for the 
selected region, organised according to the supplied experimental conditions.

## Mathematical formulation of methylation proportions

Let $N_{C,i}$ and $N_{T,i}$ denote the methylated and unmethylated read counts, 
respectively, at CpG position $i$ within a selected rDNA region.

The sequencing coverage at that position is

$$
C_i = N_{C,i} + N_{T,i}.
$$

The methylation proportion at each retained CpG position is defined as

$$
p_i =
\frac{N_{C,i}}
     {N_{C,i}+N_{T,i}}.
$$

Only CpG positions satisfying the specified minimum coverage threshold are 
included in the aggregation.

### Mean methylation proportion

The arithmetic mean methylation proportion across a region is

$$
\boxed{
\overline{p}_R =
\frac{1}{N_R}
\sum_{i=1}^{N_R}p_i
}
$$
where $N_R$ denotes the number of retained CpG positions in region $R$.

This metric assigns equal weight to every retained CpG position, regardless of 
its individual sequencing coverage. Currently there is no coverage weighted 
calculation for the methylation proportion in `myoRDNA`, if the user needs such
a metric this has to be inputted manually.

### Median methylation proportion

An alternative summary statistic is the median methylation proportion:

$$
\boxed{
\widetilde{p}_R =
\operatorname{median}
\left(p_1,p_2,\ldots,p_{N_R}\right)
}
$$
The median describes the central tendency of the CpG methylation proportions 
and is less sensitive to extreme values than the arithmetic mean. However, a 
sensible approach is to review both and determine if there is a significant 
difference. 

Regional methylation measurements can subsequently be compared between 
experimental conditions to investigate differences in the epigenetic regulation 
of rDNA.

# Differential rDNA methylation analysis

The repetitive organisation of ribosomal DNA (rDNA) presents several challenges 
for differential methylation analysis. Individual rDNA repeat units can exhibit 
heterogeneous methylation patterns, while sequence similarity between repeats 
can introduce ambiguity during read alignment.

Consequently, methylation measurements may exhibit greater variability than 
predicted by conventional binomial models. This phenomenon, known as 
*overdispersion*, can lead to underestimated standard errors and potentially 
inflated statistical significance if it is not appropriately considered.

To accommodate different levels of variability,
`myoRDNA` supports three statistical approaches:

1. Binomial logistic regression, implemented using `glm()`.
2. Quasi-binomial logistic regression, implemented using `glm()`.
3. Beta-binomial logistic regression, implemented using `glmmTMB`.

These models can be applied at two levels:

- **Site-level analysis:** Differential methylation is
  evaluated independently at individual CpG positions
  within a selected rDNA region.
- **Region-level analysis:** Methylated and unmethylated
  read counts are summed across the selected rDNA
  region before comparing experimental conditions.

All three approaches use the same logistic regression formulation, but differ 
in their assumptions about the variance of methylation measurements.

## Statistical methodology

### Binomial logistic regression

This form of regression follows the same methodology as outlined by `methylKit`
and therefore, allows users to use the widely reported and tested form of 
logistic regression. We recommend that any user of `myoRDNA` implements the
binomial method first and move to `quasi-binomial` or `beta-binomial` once they
suspect that the $p$-values are inflated. 

Binomial logistic regression models the probability of methylation as a 
function of experimental condition. Let $Y_i$ denote the number of methylated 
reads observed for sample $i$, with sequencing coverage $n_i$. 
The model assumes

$$
Y_i \sim \operatorname{Binomial}(n_i,p_i),
$$
where $p_i$ is the underlying methylation probability. For two experimental 
conditions, the model is

$$
\operatorname{logit}(p_i)
=
\log\left(\frac{p_i}{1-p_i}\right)
=
\beta_0+\beta_1 T_i.
$$
Here, $T_i=0$ represents the CONTROL condition and $T_i=1$ represents the 
CONDITION_1 condition. Multiple condition differential methylation analysis has
not been implemented yet. 

The intercept $\beta_0$ describes the log-odds of methylation in OLD samples, 
while $\beta_1$ describes the difference in log-odds between YOUNG and OLD 
samples. The model assumes that the variance of the methylated read count is

$$
\operatorname{Var}(Y_i)
=
n_i p_i(1-p_i).
$$

### Quasi-binomial logistic regression

A limitation of the conventional binomial model is its assumption that the 
variance is completely determined by the methylation probability and sequencing 
coverage.

In biological datasets, additional variability may arise from differences 
between individuals, technical variation or heterogeneous methylation across 
rDNA repeats.

Quasi-binomial logistic regression accommodates this additional variation by 
introducing a dispersion parameter, $\phi$.

The expected number of methylated reads remains

$$
\operatorname{E}(Y_i)=n_i p_i,
$$

but the variance becomes

$$
\boxed{
\operatorname{Var}(Y_i)
=
\phi n_i p_i(1-p_i)
}
$$
where $\phi$ is the dispersion parameter.

The interpretation of $\phi$ is as follows:

- $\phi=1$: The variance is consistent with the
  conventional binomial model.
- $\phi>1$: The observations exhibit overdispersion.
- $\phi<1$: The observations exhibit underdispersion.

The dispersion parameter can be estimated using the Pearson's chi-squared
statistic:

$$
\widehat{\phi}
=
\frac{\chi^{2}}{n-p}
$$
The estimated dispersion parameter is used to adjust the variance-covariance 
matrix of the regression coefficients. For overdispersed data, this generally
increases the standard errors and reduces the apparent statistical significance 
of treatment effects.

In `myoRDNA`, this model is implemented using the `quasibinomial()` family 
within R's `glm()` function.

**Important:** Reliable dispersion estimation requires sufficient residual 
degrees of freedom. Sites observed in very few samples, particularly those 
with missing coverage, may not contain enough information to estimate the 
dispersion parameter reliably. This problem is more obvious in the beta-binomial
logistic covered in the next section. Currently, in `myoRDNA` when using the 
`quasi-binomial` method there is no means of detecting how successful the 
estimation for $\hat{\phi}$ is. 

### Beta-binomial logistic regression

Beta-binomial regression provides an alternative method for modelling 
overdispersed methylation data.

Rather than assuming that each sample has a fixed underlying methylation 
probability, the beta-binomial model allows this probability to vary according 
to a beta distribution.

For sample $i$, let the underlying methylation probability be $\theta_i$, such 
that

$$
\theta_i
\sim
\operatorname{Beta}(\alpha_i,\gamma_i).
$$

Conditional on this probability, the number of methylated reads follows a 
binomial distribution:

$$
Y_i
\sim
\operatorname{BetaBinomial}
(n_i,\alpha_i,\beta_i).
$$
The expected number of methylated reads is

$$
\operatorname{E}(Y_i)=n_i p_i,
$$
In `myoRDNA`, the beta-binomial model is fitted using the 
`betabinomial(link = "logit")` family provided by `glmmTMB`.

Unlike quasi-binomial regression, beta-binomial regression specifies a complete 
likelihood, allowing maximum-likelihood estimation and likelihood-based 
statistical comparisons.

However, estimating both treatment effects and dispersion can be difficult when 
the number of biological replicates is small or methylation proportions exhibit 
little variation.

## Step 1: Site-level differential methylation

Site-level differential methylation identifies individual CpG positions that 
exhibit differences in methylation between experimental conditions.

For each genomic position, `myoRDNA` fits a separate regression model using the 
methylated and unmethylated read counts across samples.

Here, we analyse the 18S region of the previously prepared mouse WGBS dataset.

```{r differential-methylation-sites}
site_results <- diff_meth(
  meth_data = meth_rDNA_shifted,
  samples = samples,
  species = "mm39",
  num_samples = num_samples,
  min_cov = 10,
  region = "18S",
  model_type = "binomial",
  p_adjust_method = "BH",
  site = TRUE
)
```

Setting `site = TRUE` performs the differential methylation analysis 
independently at each retained CpG position.

The `min_cov = 10` argument excludes measurements with fewer than ten 
sequencing reads.

To use either of the alternative regression models, change the `model_type` 
argument: `binomial`, `quasi-binomial`, and `beta-binomial`.

Beta-binomial regression can be computationally more demanding, particularly 
when thousands of CpG positions are analysed independently. As a result of the 
algorithm requiring the use of optimisation to estimate parameters in the 
model.

## Step 2: Region-level differential methylation

Region-level differential methylation investigates differences across an entire 
annotated rDNA region, rather than fitting independent models at each CpG 
position.

For a selected region, the methylated and unmethylated read counts can be 
aggregated within each sample:

$$
M_{i,R}
=
\sum_{j \in R} N_{C,ij},
$$

$$
U_{i,R}
=
\sum_{j \in R} N_{T,ij},
$$

where $R$ denotes the selected region. The aggregated counts are then used to
estimate the association between methylation probability and experimental 
condition.

The following example performs region-level differential methylation for 18S.

```{r differential-methylation-region}
region_results <- diff_meth(
  meth_data = meth_rDNA_shifted,
  samples = samples,
  species = "mm39",
  num_samples = num_samples,
  min_cov = 10,
  region = "18S",
  model_type = "binomial",
  p_adjust_method = "BH",
  site = FALSE
)
```

Setting `site = FALSE` performs the analysis at the regional level. As with 
site-level analysis, the regression method can be changed using `model_type`.

Region-level aggregation provides a summary of methylation differences across 
the selected region. However, this approach can obscure opposing methylation 
changes at individual CpG positions and should therefore be considered 
complementary to site-level analysis.

## Step 3: Interpreting differential methylation

For the binary treatment coding between CONTROL and CONDITION 1 used in this 
example, the estimated methylation probabilities are

$$
\widehat{p}_{\mathrm{CONTROL}}
=
\frac{1}{1+\exp(-\widehat{\beta}_0)}
$$
and

$$
\widehat{p}_{\mathrm{CONDITION \ 1}}
=
\frac{1}{
1+\exp[-(\widehat{\beta}_0+\widehat{\beta}_1)]
}.
$$
An odds ratio greater than one indicates higher fitted methylation odds in 
CONDITION 1 samples, whereas an odds ratio below one indicates lower fitted 
methylation odds.

The estimated methylation difference, expressed in percentage points, is

$$
\boxed{
\Delta_{\mathrm{meth}}
=
100
\left(
\widehat{p}_{\mathrm{CONDITION \ 1}}
-
\widehat{p}_{\mathrm{CONTROL}}
\right)
}
$$
A positive value therefore indicates higher estimated methylation in CONDITION 1
samples relative to OLD samples. Statistical evidence for a treatment effect is 
assessed through the corresponding regression coefficient and its associated 
hypothesis test.

## Step 4: Multiple-testing correction

Site-level differential methylation involves testing multiple CpG positions, 
increasing the probability of obtaining false-positive results.

To account for multiple comparisons, `myoRDNA` supports p-value adjustment using
the Benjamini–Hochberg (BH) as a default procedure. However, if the user wishes 
to use a less or more conservative method this can be adjusted through the 
`p_adjust_method` to `none` for no adjustment or `bonferroni` or `holm` to name
a few. 

The BH method controls the false discovery rate (FDR), defined as the expected 
proportion of false discoveries among rejected hypotheses, under its applicable 
dependence assumptions.

The adjusted p-values are then returned to their original genomic positions.

For example, an FDR threshold of 0.05 can be used to identify CpG positions with
BH-adjusted p-values below 0.05.

The scope of the correction depends on the set of tests included. When different
rDNA regions are analysed separately, region-specific adjustments do not 
automatically control the FDR across all regions.

## Beta-binomial optimisation and model diagnostics

Beta-binomial regression estimates the regression coefficients and dispersion 
parameter by maximising the likelihood of the observed methylation counts.

Because the beta-binomial likelihood can be difficult to optimise, particularly 
with limited biological replication or extreme methylation proportions, 
numerical convergence should be assessed before interpreting the fitted 
coefficients and p-values.

### Available optimisation algorithms

The `glmmTMB` package uses the `nlminb` optimiser by default.

Alternative optimisation strategies include:

- **nlminb:** A nonlinear optimisation algorithm
  supporting gradient-based optimisation and
  parameter bounds. Its iteration and function
  evaluation limits can be increased when
  additional optimisation steps are required.

- **BFGS:** A quasi-Newton method available
  through R's `optim()` function. It approximates
  the inverse Hessian and can provide an
  alternative when the default optimiser
  encounters convergence difficulties.

- **Nelder–Mead:** A derivative-free simplex
  method available through `optim()`.
  It can be considered when gradient-based
  optimisation is problematic, although it
  may be slower or less reliable for some
  beta-binomial models.

These algorithms are alternatives for fitting the same statistical model; 
changing the optimiser does not change the model's underlying probability 
distribution.

### Assessing model convergence

Users should inspect convergence diagnostics before interpreting results from 
beta-binomial regression.

Potential problems include failure to reach convergence, a non-positive-definite
Hessian matrix, unusually large regression coefficients and poorly identified 
dispersion parameters.

These problems may arise when there is insufficient information to estimate the 
model parameters or when methylation measurements exhibit complete or 
near-complete separation.

For a fitted `glmmTMB` object,
the following diagnostics are available:

```{r beta-binomial-diagnostics, eval=FALSE}
# Inspect fitted model
summary(fit)

# Check convergence code
fit$fit$convergence

# Check whether the Hessian is positive definite
fit$sdr$pdHess

# Run additional diagnostics
glmmTMB::diagnose(fit)
```

If convergence problems arise, alternative optimisation settings may be 
considered. However, changing the optimiser or increasing iteration limits does 
not necessarily resolve parameter identifiability problems.

Results from models with unresolved convergence problems should not be 
interpreted as reliable evidence of differential methylation.

## Session information

```{r session-info}
sessionInfo()
```












